---
title: "standardizaion_modeling"
author: "Matan & Hillel & Yael"
date: "2025-07-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE}
## libraries + internal utils
library(tidyverse)    # dplyr, ggplot2, purrr
library(glmnet)       # L1 logistic
library(ranger)       # random forest
library(doParallel)   # parallel bootstrap
library(foreach)
library(knitr)        # nice tables

source("utils/data_pipelines_utils.R")          # impute_df()
source("utils/plot_utils.R")                  # optional plotting
source("utils/causal_utils_standardization.R")
```

#### Data Loading & Basic Preprocessing

```{r}
base_df <- read_csv("data/full_cohort_data.csv", show_col_types = FALSE)

# Treatment & outcome
treatment <- "aline_flg"
outcome   <- "day_28_flg"

# Covariate blocks (identical to IPTW notebook)
individual_properties <- c(
  "age", "gender_num", "weight_first", "bmi"
)

severity_scores <- c(
  "sapsi_first", "sofa_first"
)

vital_signs <- c(
  "map_1st", "hr_1st", "temp_1st", "spo2_1st"
)

lab_results <- c(
  "wbc_first", "hgb_first", "platelet_first", "sodium_first",
  "potassium_first", "tco2_first", "chloride_first", "bun_first",
  "creatinine_first", "po2_first", "pco2_first"
)

first_day <- c("iv_day_1")

comorbidities <- c(
  "chf_flg", "afib_flg", "renal_flg", "liver_flg", "copd_flg",
  "cad_flg", "stroke_flg", "mal_flg", "resp_flg"
)

admission_properties <- c("service_unit", "day_icu_intime", "hour_icu_intime")

confounders <- c(
  individual_properties, severity_scores, vital_signs, lab_results,
  first_day, comorbidities, admission_properties
)

```

We apply same stratification we used in the IPTW notebook to handle positivity violation of the SOFA first variable and remove the bmi covariate.

```{r}
clean_df <- base_df %>% select(-bmi)
clean_confounders <- confounders[!confounders %in% "bmi"]

final_base_df <- clean_df %>%
  mutate(sofa_first_cat = case_when(
    sofa_first < 2.5 ~ "0-2.5",
    sofa_first < 5   ~ "2.5-5",
    sofa_first < 7.5 ~ "5-7.5",
    sofa_first < 10  ~ "7.5-10",
    TRUE             ~ "10+"
  )) %>%
  select(-sofa_first)

final_confounders <- clean_confounders[!clean_confounders %in% "sofa_first"] %>%
  append("sofa_first_cat")

```

We divide the data into 3 pipelines based on different imputation methods.

```{r}
df_no_na <- final_base_df %>%
  select(all_of(c(treatment, outcome, final_confounders))) %>%
  filter(complete.cases(.))

df_mean_imputed <- final_base_df %>%
  select(all_of(c(treatment, outcome, final_confounders))) %>%
  impute_df(method = "mean")

df_median_imputed <- final_base_df %>%
  select(all_of(c(treatment, outcome, final_confounders))) %>%
  impute_df(method = "median")

```

#### Pipeline Definition

We run 9 pipelines (3 imputation methods x 3 models).\
In each pipeline we run bootstrap estimation by fitting the model and calculating the ATE using standardization. We also present plots of the ATE empirical distribution and the model's predicted probabilities conditioned on the treatment assignment.

```{r}
run_pipeline <- function(df, dataset_name, model_type) {

  set.seed(42)
  cat("\n----", dataset_name, "|", model_type, "----\n")

  ## 1) fit outcome model
  fit <- model_outcome(
    df, final_confounders, treatment, outcome, model_type
  )

  ## 2) counterfactual predictions
  preds <- predict_counterfactual(
    fit, df, final_confounders, treatment, outcome
  )

  ## 3) point estimate
  est <- standardization_ate(preds)

  ## 4) bootstrap variance + ATE histogram  (prints automatically)
  boot <- bootstrap_standardization_ate(
    df, final_confounders, treatment, outcome,
    model_type   = model_type,
    n_bootstrap  = 200,
    plot_title   = sprintf("Bootstrap ATE – %s | %s",
                           dataset_name, model_type)
  )

  ## 5) predicted-risk histogram (observed treatment group assignment)
  observed_pred <- ifelse(df[[treatment]] == 1, preds$y1_hat, preds$y0_hat)

  plot_predicted_risk_hist(
  df, observed_pred,
  group_col = outcome,                           # <- NEW ARG
  plot_title = sprintf("Predicted risk by outcome – %s | %s",
                       dataset_name, model_type)
)

  ## 6) return tidy one-row tibble
  tibble(
    dataset  = dataset_name,
    model    = model_type,
    mu1      = est$mu1,
    mu0      = est$mu0,
    ATE      = est$ATE,
    CI_lower = boot$ci[1],
    CI_upper = boot$ci[2],
    SD       = boot$sd
  )
}

```

#### Pipeline Running

```{r}
datasets <- list(
  no_na          = df_no_na,
  mean_imputed   = df_mean_imputed,
  median_imputed = df_median_imputed
)

model_types <- c("logreg", "logreg_l1", "rf")
```

```{r}
results_no_na <- purrr::map_dfr(
  model_types,
  ~ run_pipeline(df_no_na, "no_na", .x)
)
```

```{r}
results_mean <- purrr::map_dfr(
  model_types,
  ~ run_pipeline(df_mean_imputed, "mean_imputed", .x)
)
```

```{r}
results_median <- purrr::map_dfr(
  model_types,
  ~ run_pipeline(df_median_imputed, "median_imputed", .x)
)
```

#### Final Results Comparison

```{r}
results_tbl <- bind_rows(results_no_na, results_mean, results_median)

kable(results_tbl,
      digits = 3,
      caption = "Standardisation ATE (bootstrap 95% CI, 200 draws)")

```
