---
title: "basic_eda"
author: "Matan & Hillel & Yael"
date: "2025-07-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### File Description

This file performs exploratory analysis of the data: it plots 28‑day mortality by IAC treatment, severity‑score quartiles, comorbidities, vitals, labs and demographics; assesses variable missing values; and shows density overlays comparing raw, mean‑ and median‑imputed values, supplying counts, risks and CIs throughout.

```{r}
library(tidyverse)
source("utils/data_pipelines_utils.R")
source("utils/plot_utils.R")
source("utils/ate_utils_iptw.R")
full_df <- read.csv("data/full_cohort_data.csv")
```

```{r}
col_alive <- "#8DBFAF"  # teal
col_death <- "#D8A3A8"  # coral
col_point <- "#D8A3A8"
```

### Outcome by Treatment

```{r}
## Counts bar chart (Alive / Death)
counts_plot <- full_df %>% 
  group_by(aline_flg, day_28_flg) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  mutate(outcome = factor(day_28_flg, levels = c(0, 1),
                          labels  = c("Alive", "Death")),
         tx      = factor(aline_flg, levels = c(0, 1),
                          labels  = c("No IAC", "IAC"))) %>% 
  ggplot(aes(tx, n, fill = outcome)) +
  geom_col(position = position_dodge(0.7),
           width = .6, colour = "grey30") +
  scale_y_continuous(expand = expansion(mult = c(0, .05))) +
  scale_fill_manual(values = c("Alive" = col_alive,
                               "Death" = col_death),
                    name = "28-day\noutcome") +
  labs(title = "28-Day Mortality (outcome), by IAC (treatment)",
       x = NULL, y = "Number of patients")

counts_plot
```

```{r}
## Risk dot + CI plot
risk_tbl <- full_df %>% 
  group_by(tx = factor(aline_flg, levels = c(0, 1),
                       labels = c("No IAC", "IAC"))) %>% 
  summarise(
    deaths = sum(day_28_flg == 1),
    n      = n(),
    risk   = deaths / n,
    .groups = "drop"
  ) %>% 
  mutate(
    se  = sqrt(risk * (1 - risk) / n),
    lwr = pmax(0, risk - 1.96 * se),
    upr = pmin(1, risk + 1.96 * se)
  )

risk_plot <- ggplot(risk_tbl, aes(tx, risk, colour = tx)) +
  geom_point(size = 5, colour = "#D8A3A8") +
  geom_errorbar(aes(ymin = lwr, ymax = upr),
                width = .15, linewidth = 0.8, colour = "#D8A3A8") +
  scale_y_continuous(labels = scales::percent_format(1),
                     limits = c(0, 0.30),
                     expand = expansion(mult = c(0, .05))) +
  labs(title = "28-Day Mortality (outcome) Risk, by IAC (treatment)",
       subtitle = "Estimated Wald 95% CI",
       x = NULL, y = "Observed risk (%)")

risk_tbl
```

### Severity-Score Quartile Trends

```{r}
severity_scores <- c("sapsi_first", "sofa_first")

for (score in severity_scores) {
  
  # quartiles
  q_tbl <- full_df %>% 
    filter(!is.na(.data[[score]])) %>% 
    mutate(quartile = ntile(.data[[score]], 4L)) %>% 
    group_by(quartile) %>% 
    summarise(
      deaths = sum(day_28_flg == 1),
      n      = n(),
      risk   = deaths / n,
      .groups = "drop"
    ) %>% 
    mutate(
      se  = sqrt(risk * (1 - risk) / n),
      lwr = pmax(0, risk - 1.96 * se),
      upr = pmin(1, risk + 1.96 * se),
      quartile = factor(quartile,
                        labels = c("Lowest","Mid-Low","Mid-High","Highest"))
    )
  
  p <- ggplot(q_tbl, aes(quartile, risk, group = 1)) +
    geom_point(size = 4, colour = col_point) +
    geom_errorbar(aes(ymin = lwr, ymax = upr),
                  width = .15, colour = col_point, linewidth = 0.8) +
    geom_line(colour = "grey50", linewidth = 0.6, linetype = 2) +
    scale_y_continuous(labels = scales::percent_format(1),
                       limits = c(0, 0.35),
                       expand = expansion(mult = c(0, .05))) +
    labs(title = paste("28-day Mortality (outcome) Risk, by", toupper(score), "Quartile"),
         x = paste(toupper(score), "Quartile"),
         y = "Observed Risk (%)")
  
  print(p)
}
```

### SAPS I vs SOFA Scatter

```{r}
scatter_df <- full_df %>% 
  filter(!is.na(sapsi_first), !is.na(sofa_first)) %>% 
  mutate(outcome = factor(day_28_flg, 0:1, c("Alive","Death")))

scatter_plot <- ggplot(scatter_df,
                       aes(sapsi_first, sofa_first, colour = outcome)) +
  geom_point(alpha = 0.40, size = 2) +
  scale_colour_manual(values = c("Alive" = col_alive,
                                 "Death" = col_death),
                      name = "28-day\noutcome") +
  labs(title = "Joint distribution of SAPS I and SOFA",
       x = "SAPS I (first 24 h)",
       y = "SOFA (first 24 h)") +
  coord_cartesian(expand = 0.02)

scatter_df
```

### Comorbidity risk plots

```{r}
comorbidities <- c(
  "chf_flg", "afib_flg", "renal_flg", "liver_flg", "copd_flg",
  "cad_flg", "stroke_flg", "mal_flg", "resp_flg"
)

for (var in comorbidities) {
  
  tbl <- full_df %>% 
    filter(!is.na(.data[[var]])) %>% 
    mutate(status = factor(.data[[var]], 0:1, c("No","Yes"))) %>% 
    group_by(status) %>% 
    summarise(
      deaths = sum(day_28_flg == 1),
      n      = n(),
      risk   = deaths / n,
      .groups = "drop"
    ) %>% 
    mutate(
      se  = sqrt(risk * (1 - risk) / n),
      lwr = pmax(0, risk - 1.96 * se),
      upr = pmin(1, risk + 1.96 * se)
    )
  
  p <- ggplot(tbl, aes(status, risk, group = 1)) +
    geom_point(size = 4, colour = col_point) +
    geom_errorbar(aes(ymin = lwr, ymax = upr),
                  width = .15, colour = col_point, linewidth = 0.8) +
    geom_line(colour = "grey50", linewidth = 0.6, linetype = 2) +
    scale_y_continuous(labels = scales::percent_format(1),
                       limits = c(0, 0.5),
                       expand = expansion(mult = c(0, .05))) +
    labs(title = paste("28-day Mortality (outcome) Risk, by", toupper(var)),
         x = paste(toupper(var), "(Yes / No)"),
         y = "Observed Risk (%)")
  
  print(p)
}
```

### Continuous Vital Signs & Lab Results

```{r}
vital_signs <- c("map_1st", "hr_1st",  "temp_1st", "spo2_1st")
lab_results <- c(
  "wbc_first", "hgb_first", "platelet_first", "sodium_first", "potassium_first",
  "tco2_first", "chloride_first", "bun_first", "creatinine_first",
  "po2_first", "pco2_first"
)

cont_vars <- c(vital_signs, lab_results)

for (var in cont_vars) {
  
  df <- full_df %>% 
    filter(!is.na(.data[[var]])) %>% 
    mutate(outcome = factor(day_28_flg, 0:1, c("Alive","Death")))
  
  p <- ggplot(df, aes(outcome, .data[[var]], fill = outcome)) +
    geom_boxplot(
      alpha         = 0.35,
      colour        = "grey30",
      width         = 0.55,
      outlier.shape = 16,       # solid circles for outliers
      outlier.size  = 1.4,
      outlier.alpha = 0.4
      # coef = 1.5 is the default → Tukey whiskers, horizontal caps included
    ) +
    scale_fill_manual(
      values = c("Alive" = col_alive,
                 "Death" = col_death),
      guide  = "none"
    ) +
    labs(
      title = paste(toupper(var), "distribution by 28-day outcome"),
      x     = NULL,
      y     = toupper(var)
    )
  
  print(p)
}
```

### Individual-Level Properties

```{r}
individual_properties <- c("age", "gender_num", "weight_first", "bmi")

for (var in individual_properties) {
  
  if (var == "gender_num") {
    df <- full_df %>% 
      filter(!is.na(gender_num)) %>% 
      mutate(
        gender  = factor(gender_num, 0:1, c("Female", "Male")),
        outcome = factor(day_28_flg, 0:1, c("Alive","Death"))
      )
    
    p <- ggplot(df, aes(gender, fill = outcome)) +
      geom_bar(position = "fill", colour = "grey30") +
      scale_y_continuous(labels = scales::percent_format()) +
      scale_fill_manual(values = c("Alive" = col_alive,
                                   "Death" = col_death),
                        name   = "28-day\noutcome") +
      labs(
        title = "Gender distribution by 28-day outcome",
        x     = NULL,
        y     = "Within-gender %"
      )
    
  } else {
    df <- full_df %>% 
      filter(!is.na(.data[[var]])) %>% 
      mutate(outcome = factor(day_28_flg, 0:1, c("Alive","Death")))
    
    p <- ggplot(df, aes(.data[[var]], fill = outcome, colour = outcome)) +
      geom_density(alpha = 0.35, linewidth = 0.8, adjust = 1.1) +
      scale_fill_manual(values = c("Alive" = col_alive,
                                   "Death" = col_death),
                        guide = "none") +
      scale_colour_manual(values = c("Alive" = col_alive,
                                     "Death" = col_death),
                          guide = "none") +
      labs(
        title = paste(toupper(var), "density by 28-day Mortality (outcome)"),
        x     = toupper(var),
        y     = "Density"
      )
  }
  
  print(p)
}
```

### Missing Values Interpretation

```{r}
missing_summary <- full_df %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
  pivot_longer(everything(),
               names_to  = "variable",
               values_to = "pct_missing") %>%
  filter(pct_missing > 0) %>%
  arrange(desc(pct_missing))

p_missing <- ggplot(missing_summary,
                    aes(x = pct_missing,
                        y = reorder(variable, pct_missing))) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = sprintf("%.1f%%", pct_missing)),
            hjust = -0.05, size = 3) +
  labs(title = "Percent missing per variable",
       x = "% missing", y = NULL) +
  coord_cartesian(xlim = c(0, max(missing_summary$pct_missing) * 1.15))

print(p_missing)

```

We will exclude BMI and re-plot the missing values by variable:

```{r}
missing_summary_nobmi <- full_df %>% 
  select(-bmi) %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>% 
  pivot_longer(everything(),
               names_to  = "variable",
               values_to = "pct_missing") %>% 
  filter(pct_missing > 0) %>% 
  arrange(desc(pct_missing))

p_missing_nobmi <- ggplot(missing_summary_nobmi,
                          aes(x = pct_missing,
                              y = reorder(variable, pct_missing))) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = sprintf("%.1f%%", pct_missing)),
            hjust = -0.05, size = 3) +
  labs(title = "Percent missing per variable (BMI removed)",
       x = "% missing", y = NULL) +
  coord_cartesian(xlim = c(0, max(missing_summary_nobmi$pct_missing) * 1.15))

p_missing_nobmi
```

### Imputation Methods Diagnostic

```{r}
num_cols <- names(which(sapply(full_df, is.numeric)))

df_mean_imputed <- full_df %>% 
  select(-bmi) %>%
  mutate(across(all_of(setdiff(num_cols, "bmi")), 
                ~ ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x)))

df_median_imputed <- full_df %>% 
  mutate(across(all_of(setdiff(num_cols, "bmi")),
                ~ ifelse(is.na(.x), median(.x, na.rm = TRUE), .x)))
```

```{r}
missing_vars <- intersect(missing_summary$variable, num_cols)

for (v in missing_vars) {

  combined <- bind_rows(
    No_NA  = full_df,
    Mean   = df_mean_imputed,
    Median = df_median_imputed,
    .id    = "set"
  ) %>%
    select(set, !!sym(v)) %>%
    filter(!is.na(.data[[v]]))

  p <- ggplot(combined,
              aes(.data[[v]], colour = set, fill = set)) +
    geom_density(alpha = 0.20, adjust = 1.1) +
    scale_colour_manual(values = c(No_NA = "grey30",
                                   Mean  = "#1F78B4",
                                   Median = "#33A02C"),
                        name = NULL) +
    scale_fill_manual(values = c(No_NA = NA,
                                 Mean  = "#1F78B4",
                                 Median = "#33A02C"),
                      guide = "none") +
    labs(title = paste("Observed vs imputed:", v),
         x = v, y = "Density")

  print(p)   # shows in the active graphics device
}
```
