---
title: "Propensity Scores & IPTW"
author: "Matan, Hillel & Yael"
date: "2025-06-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE}
# libraries, internal imports and constants
library(tidyverse)
source("utils/data_pipelines_utils.R")
source("utils/plot_utils.R")
source("utils/causal_utils.R")
```

```{r}
base_df <- read.csv("data/full_cohort_data.csv")

individual_properties <- c(
  "age", "gender_num", "weight_first", "bmi"
  )

sevirity_scores <- c(
  "sapsi_first", "sofa_first"
)

vital_signs <- c(
  "map_1st", "hr_1st", "temp_1st", "spo2_1st"
)

lab_results <- c(
  "wbc_first", "hgb_first", "platelet_first", "sodium_first", "potassium_first",
  "tco2_first", "chloride_first", "bun_first", "creatinine_first", "po2_first",
  "pco2_first"
)

first_day <- c("iv_day_1")

comorbidities <- c(
  "chf_flg", "afib_flg", "renal_flg", "liver_flg", "copd_flg",
  "cad_flg", "stroke_flg", "mal_flg", "resp_flg"
)

# stratify hour icu into bins for categorical variable
admission_properties <- c("service_unit", "day_icu_intime", "hour_icu_intime")

treatment <- c("aline_flg")
outcome <- c("day_28_flg")

confounders <- c(
  individual_properties, sevirity_scores, vital_signs, lab_results,
  first_day, comorbidities, admission_properties
)
```

## Missing Values EDA

```{r}
# Percentage of rows with at least one missing value
total_rows        <- nrow(base_df)
rows_with_missing <- sum(!complete.cases(base_df))
pct_rows_missing  <- rows_with_missing / total_rows * 100
cat(sprintf("Rows with ≥1 missing: %d / %d (%.2f%%)\n", 
            rows_with_missing, total_rows, pct_rows_missing))

# Percentage of missing values per column
col_pct_missing <- sapply(base_df, function(col) mean(is.na(col)) * 100)

missing_summary <- data.frame(
  variable    = names(col_pct_missing),
  pct_missing = col_pct_missing,
  row.names   = NULL
)

print(missing_summary %>% filter(pct_missing > 0))
```

```{r}
clean_df <- base_df %>%
  select(-bmi)

clean_confounders <- confounders[!confounders %in% "bmi"]
```

## Validating positivity across all confounders

```{r}
positivity_check_histograms(clean_df, treatment, clean_confounders)

```

Suspected positivity issues:\
po2_first - we deny this\
sofa_first - divide to the intervals: [0, 2,5) [2.5, 5) [5, 7.5) [7.5, 10} [10, \\inf)

## Dividing sofa_first

```{r}
final_base_df <- clean_df %>%
  mutate(sofa_first_cat = case_when(
    sofa_first < 2.5 ~ "0-2.5",
    sofa_first < 5   ~ "2.5-5",
    sofa_first < 7.5 ~ "5-7.5",
    sofa_first < 10  ~ "7.5-10",
    TRUE             ~ "10+"
  )) %>%
  select(-sofa_first)

# change sofa first to sofa_first_cat in the confounders on clean confounders
final_confounders <- clean_confounders[!clean_confounders %in% "sofa_first"]
final_confounders <- c(final_confounders, "sofa_first_cat")

positivity_check_histograms(final_base_df, treatment, final_confounders)
```

```{r}
# Final datasets based on the final_base_df

# Removed all NA values
df_no_na <- final_base_df %>%
  filter(complete.cases(final_base_df))

# mean imputation
df_mean_imputed <- impute_df(final_base_df, method = "mean")

# median imputation
df_median_imputed <- impute_df(final_base_df, method = "median")
```

## Estimate PS for 3 datasets

```{r}
# Estimate glm models for each ones without lasso
logreg_no_na_no_lasso <- model_ps_score(df_no_na, final_confounders, treatment, lasso = FALSE)
logreg_mean_imputed_no_lasso <- model_ps_score(df_mean_imputed, final_confounders, treatment, lasso = FALSE)
logreg_median_imputed_no_lasso <- model_ps_score(df_median_imputed, final_confounders, treatment, lasso = FALSE)

# Estimate glm models for each ones with lasso
logreg_no_na_lasso <- model_ps_score(df_no_na, final_confounders, treatment, lasso = TRUE)
logreg_mean_imputed_lasso <- model_ps_score(df_mean_imputed, final_confounders, treatment, lasso = TRUE)
logreg_median_imputed_lasso <- model_ps_score(df_median_imputed, final_confounders, treatment, lasso = TRUE)
```

```{r}
# extract propensity‐scores using extract_ps()
ps_no_na_no_lasso         <- extract_ps(logreg_no_na_no_lasso)
ps_mean_imputed_no_lasso  <- extract_ps(logreg_mean_imputed_no_lasso)
ps_median_imputed_no_lasso<- extract_ps(logreg_median_imputed_no_lasso)

ps_no_na_lasso            <- extract_ps(logreg_no_na_lasso,
                                        df_no_na,
                                        final_confounders,
                                        treatment)
ps_mean_imputed_lasso     <- extract_ps(logreg_mean_imputed_lasso,
                                        df_mean_imputed,
                                        final_confounders,
                                        treatment)
ps_median_imputed_lasso   <- extract_ps(logreg_median_imputed_lasso,
                                        df_median_imputed,
                                        final_confounders,
                                        treatment)

```

```{r}
# Plot propensity scores for each one of the 6 models

# No lasso models
plot_ps_hist(df_no_na, ps_no_na_no_lasso, treatment, plot_title="PS No NA | No Lasso")
plot_ps_hist(df_mean_imputed, ps_mean_imputed_no_lasso, treatment, plot_title="PS Mean Imputation | No Lasso")
plot_ps_hist(df_median_imputed, ps_median_imputed_no_lasso, treatment,
             plot_title="PS Median Imputation | No Lasso")

# With lasso models
plot_ps_hist(df_no_na, ps_no_na_lasso, treatment, plot_title="PS No NA | Lasso")
plot_ps_hist(df_mean_imputed, ps_mean_imputed_lasso, treatment, 
             plot_title="PS Mean Imputation | Lasso")
plot_ps_hist(df_median_imputed, ps_median_imputed_lasso, treatment,
             plot_title="PS Median Imputation | Lasso")
```

## Estimating ATE on 6 models

#### IPTW Weights Un-Clipped computation + histograms

```{r}
# compute iptw weights for each of the 6 models, un - clipped
w_no_na_not_clipped <- iptw_weights(
  df            = df_no_na,
  ps_vec        = ps_no_na_no_lasso,
  treatment_col = treatment,
  stabilized    = TRUE,
  clip = FALSE
)

w_mean_imputed_not_clipped <- iptw_weights(
  df            = df_mean_imputed,
  ps_vec        = ps_mean_imputed_no_lasso,
  treatment_col = treatment,
  stabilized    = TRUE,
  clip = FALSE
)

w_median_imputed_not_clipped <- iptw_weights(
  df            = df_median_imputed,
  ps_vec        = ps_median_imputed_no_lasso,
  treatment_col = treatment,
  stabilized    = TRUE,
  clip = FALSE
)

w_no_na_lasso_not_clipped <- iptw_weights(
  df            = df_no_na,
  ps_vec        = ps_no_na_lasso,
  treatment_col = treatment,
  stabilized    = TRUE,
  clip = FALSE
)

w_mean_imputed_lasso_not_clipped <- iptw_weights(
  df            = df_mean_imputed,
  ps_vec        = ps_mean_imputed_lasso,
  treatment_col = treatment,
  stabilized    = TRUE,
  clip = FALSE
)

w_median_imputed_lasso_not_clipped <- iptw_weights(
  df            = df_median_imputed,
  ps_vec        = ps_median_imputed_lasso,
  treatment_col = treatment,
  stabilized    = TRUE,
  clip = FALSE
)
```

```{r}
# plot iptw weights histograms for all 6 models based on un - clipped
plot_iptw_weights(
  df            = df_no_na,
  weights       = w_no_na_not_clipped,
  treatment_col = treatment,
  bins          = 30,
  plot_title    = "IPTW Weights (No NA, Stabilized, Not Clipped)"
)

plot_iptw_weights(
  df            = df_mean_imputed,
  weights       = w_mean_imputed_not_clipped,
  treatment_col = treatment,
  bins          = 30,
  plot_title    = "IPTW Weights (Mean Imputation, Stabilized, Not Clipped)"
)

plot_iptw_weights(
  df            = df_median_imputed,
  weights       = w_median_imputed_not_clipped,
  treatment_col = treatment,
  bins          = 30,
  plot_title    = "IPTW Weights (Median Imputation, Stabilized, Not Clipped)"
)

plot_iptw_weights(
  df            = df_no_na,
  weights       = w_no_na_lasso_not_clipped,
  treatment_col = treatment,
  bins          = 30,
  plot_title    = "IPTW Weights (No NA, Lasso, Stabilized, Not Clipped)"
)

plot_iptw_weights(
  df            = df_mean_imputed,
  weights       = w_mean_imputed_lasso_not_clipped,
  treatment_col = treatment,
  bins          = 30,
  plot_title    = "IPTW Weights (Mean Imputation, Lasso, Stabilized, Not Clipped)"
)

plot_iptw_weights(
  df            = df_median_imputed,
  weights       = w_median_imputed_lasso_not_clipped,
  treatment_col = treatment,
  bins          = 30,
  plot_title    = "IPTW Weights (Median Imputation, Lasso, Stabilized, Not Clipped)"
)
```

#### IPTW Weights Clipped computation + histograms

```{r}
# compute iptw weights for each of the 6 models, clipped

w_no_na <- iptw_weights(
  df            = df_no_na,
  ps_vec        = ps_no_na_no_lasso,
  treatment_col = treatment,
  stabilized    = TRUE,
  clip = TRUE
)

w_mean_imputed <- iptw_weights(
  df            = df_mean_imputed,
  ps_vec        = ps_mean_imputed_no_lasso,
  treatment_col = treatment,
  stabilized    = TRUE,
  clip = TRUE
)

w_median_imputed <- iptw_weights(
  df            = df_median_imputed,
  ps_vec        = ps_median_imputed_no_lasso,
  treatment_col = treatment,
  stabilized    = TRUE,
  clip = TRUE
)

w_no_na_lasso <- iptw_weights(
  df            = df_no_na,
  ps_vec        = ps_no_na_lasso,
  treatment_col = treatment,
  stabilized    = TRUE,
  clip = TRUE
)

w_mean_imputed_lasso <- iptw_weights(
  df            = df_mean_imputed,
  ps_vec        = ps_mean_imputed_lasso,
  treatment_col = treatment,
  stabilized    = TRUE,
  clip = TRUE
)


w_median_imputed_lasso <- iptw_weights(
  df            = df_median_imputed,
  ps_vec        = ps_median_imputed_lasso,
  treatment_col = treatment,
  stabilized    = TRUE,
  clip = TRUE
)


```

```{r}
# plot iptw weights histograms for all 6 models based on clipped

plot_iptw_weights(
  df            = df_no_na,
  weights       = w_no_na,
  treatment_col = treatment,
  bins          = 30,
  plot_title    = "IPTW Weights (No NA, Stabilized)"
)

plot_iptw_weights(
  df            = df_mean_imputed,
  weights       = w_mean_imputed,
  treatment_col = treatment,
  bins          = 30,
  plot_title    = "IPTW Weights (Mean Imputation, Stabilized)"
)

plot_iptw_weights(
  df            = df_median_imputed,
  weights       = w_median_imputed,
  treatment_col = treatment,
  bins          = 30,
  plot_title    = "IPTW Weights (Median Imputation, Stabilized)"
)

plot_iptw_weights(
  df            = df_no_na,
  weights       = w_no_na_lasso,
  treatment_col = treatment,
  bins          = 30,
  plot_title    = "IPTW Weights (No NA, Lasso, Stabilized)"
)

plot_iptw_weights(
  df            = df_mean_imputed,
  weights       = w_mean_imputed_lasso,
  treatment_col = treatment,
  bins          = 30,
  plot_title    = "IPTW Weights (Mean Imputation, Lasso, Stabilized)"
)

plot_iptw_weights(
  df            = df_median_imputed,
  weights       = w_median_imputed_lasso,
  treatment_col = treatment,
  bins          = 30,
  plot_title    = "IPTW Weights (Median Imputation, Lasso, Stabilized)"
)
```

```{r}
# Estimate ATE For all models with clipping and stabilizing
ate_no_na <- iptw_ate(
  df            = df_no_na,
  outcome_col   = outcome,
  treatment_col = treatment,
  ps_vec        = ps_no_na_no_lasso,
  stabilized    = TRUE,
  clip          = TRUE
)

ate_mean_imputed <- iptw_ate(
  df            = df_mean_imputed,
  outcome_col   = outcome,
  treatment_col = treatment,
  ps_vec        = ps_mean_imputed_no_lasso,
  stabilized    = TRUE,
  clip          = TRUE
)

ate_median_imputed <- iptw_ate(
  df            = df_median_imputed,
  outcome_col   = outcome,
  treatment_col = treatment,
  ps_vec        = ps_median_imputed_no_lasso,
  stabilized    = TRUE,
  clip          = TRUE
)

ate_no_na_lasso <- iptw_ate(
  df            = df_no_na,
  outcome_col   = outcome,
  treatment_col = treatment,
  ps_vec        = ps_no_na_lasso,
  stabilized    = TRUE,
  clip          = TRUE
)

ate_mean_imputed_lasso <- iptw_ate(
  df            = df_mean_imputed,
  outcome_col   = outcome,
  treatment_col = treatment,
  ps_vec        = ps_mean_imputed_lasso,
  stabilized    = TRUE,
  clip          = TRUE
)

ate_median_imputed_lasso <- iptw_ate(
  df            = df_median_imputed,
  outcome_col   = outcome,
  treatment_col = treatment,
  ps_vec        = ps_median_imputed_lasso,
  stabilized    = TRUE,
  clip          = TRUE
)

# Print ATE results for each model
ate_results <- list(
  no_na                = ate_no_na,
  mean_imputed         = ate_mean_imputed,
  median_imputed       = ate_median_imputed,
  no_na_lasso          = ate_no_na_lasso,
  mean_imputed_lasso   = ate_mean_imputed_lasso,
  median_imputed_lasso = ate_median_imputed_lasso
)

for (model_name in names(ate_results)) {
  ate <- ate_results[[model_name]]
  cat(sprintf("Model: %s\n", model_name))
  cat(sprintf("Mean Treated: %.3f\n", ate$mean_treated))
  cat(sprintf("Mean Control: %.3f\n", ate$mean_control))
  cat(sprintf("ATE: %.3f\n", ate$ATE))
  cat("\n")
}
```

## Variance Estimation

```{r}
res1 <- bootstrap_iptw_ate(
  df            = df_no_na,
  confounders   = final_confounders,
  treatment_col = treatment,
  outcome_col   = outcome,
  lasso         = FALSE,
  plot_title    = "Bootstrap IPTW-ATE (No NA, No Lasso)"
)
```

```{r}
res2 <- bootstrap_iptw_ate(
  df            = df_mean_imputed,
  confounders   = final_confounders,
  treatment_col = treatment,
  outcome_col   = outcome,
  lasso         = FALSE,
  plot_title    = "Bootstrap IPTW-ATE (Mean Imputation, No Lasso)"
)
```

```{r}
res3 <- bootstrap_iptw_ate(
  df            = df_median_imputed,
  confounders   = final_confounders,
  treatment_col = treatment,
  outcome_col   = outcome,
  lasso         = FALSE,
  plot_title    = "Bootstrap IPTW-ATE (Median Imputation, No Lasso)"
)
```

```{r}
res4 <- bootstrap_iptw_ate(
  df            = df_no_na,
  confounders   = final_confounders,
  treatment_col = treatment,
  outcome_col   = outcome,
  lasso         = TRUE,
  plot_title    = "Bootstrap IPTW-ATE (No NA, Lasso)"
)
```

```{r}
res5 <- bootstrap_iptw_ate(
  df            = df_mean_imputed,
  confounders   = final_confounders,
  treatment_col = treatment,
  outcome_col   = outcome,
  lasso         = TRUE,
  plot_title    = "Bootstrap IPTW-ATE (Mean Imputation, Lasso)"
)
```

```{r}
res6 <- bootstrap_iptw_ate(
  df            = df_median_imputed,
  confounders   = final_confounders,
  treatment_col = treatment,
  outcome_col   = outcome,
  lasso         = TRUE,
  plot_title    = "Bootstrap IPTW-ATE (Median Imputation, Lasso)"
)
```

## Analyzing Extreme PS Values

To analyze the extreme PS values, we will create a stable row-level identifier across datasets, extract the propensity scores, and identify the extremes in each model’s PS distribution. 
We will then select the top and bottom percentiles (90th and 10th) of PS values across all models and compare these extreme groups on each baseline variable.

```{r}
# Add a stable row-level identifier ad-hoc for identify extreme PS values across datasets

for (nm in c("df_no_na", "df_mean_imputed", "df_median_imputed")) {
  tmp <- get(nm)
  tmp$row_id <- as.integer(row.names(tmp))   # preserves original index
  assign(nm, tmp, envir = .GlobalEnv)
}

id_var  <- "row_id"
treat   <- "aline_flag"
```

```{r}
# Create a list of data frames with propensity scores
ps_dfs <- list(
  no_na_no_lasso      = tibble(row_id = df_no_na$row_id,
                               no_na_no_lasso = ps_no_na_no_lasso),
  mean_imp_no_lasso   = tibble(row_id = df_mean_imputed$row_id,
                               mean_imp_no_lasso = ps_mean_imputed_no_lasso),
  median_imp_no_lasso = tibble(row_id = df_median_imputed$row_id,
                               median_imp_no_lasso = ps_median_imputed_no_lasso),
  no_na_lasso         = tibble(row_id = df_no_na$row_id,
                               no_na_lasso = ps_no_na_lasso),
  mean_imp_lasso      = tibble(row_id = df_mean_imputed$row_id,
                               mean_imp_lasso = ps_mean_imputed_lasso),
  median_imp_lasso    = tibble(row_id = df_median_imputed$row_id,
                               median_imp_lasso = ps_median_imputed_lasso)
)

# Full-join one after another → wide table, 1 row per row_id
master <- reduce(ps_dfs, full_join, by = "row_id")
```

```{r}
# Define the cut-offs
top_q    <- 0.9
bottom_q <- 0.1

# Grab IDs in each tail for every PS column
top_ids <- lapply(master[-1], \(ps)
  master$row_id[ ps >= quantile(ps, probs = top_q,    na.rm = TRUE) ])

bot_ids <- lapply(master[-1], \(ps)
  master$row_id[ ps <= quantile(ps, probs = bottom_q, na.rm = TRUE) ])

# Keep only patients that appear in all tail-sets
common_top <- Reduce(intersect, top_ids)
common_bot <- Reduce(intersect, bot_ids)

length(common_top)   # how many in top tail everywhere?
length(common_bot)   # how many in bottom tail everywhere?

```

```{r}
# Create a table with quantiles for each model
q_tbl <- bind_rows(
  lapply(names(master[-1]), function(col) {
    tibble(
      model       = col,
      q_bottom10  = quantile(master[[col]], probs = bottom_q, na.rm = TRUE),
      q_top90     = quantile(master[[col]], probs = top_q,    na.rm = TRUE)
    )
  })
)

print(q_tbl, n = nrow(q_tbl))
```


```{r}
# Flag extremes in each dataset
flag_extremes <- function(df) {
  df %>% mutate(
    extreme_group = case_when(
      row_id %in% common_top ~ "Top-tail (all models)",
      row_id %in% common_bot ~ "Bottom-tail (all models)",
      TRUE                   ~ "Other"
    )
  )
}

df_no_na          <- flag_extremes(df_no_na)
df_mean_imputed   <- flag_extremes(df_mean_imputed)
df_median_imputed <- flag_extremes(df_median_imputed)

```

```{r}
# Create a table comparing extreme groups by each variable
library(tableone)

vars_to_show <- setdiff(names(df_no_na),
                        c(id_var, treat, "extreme_group"))

tab <- CreateTableOne(
  vars       = vars_to_show,
  strata     = "extreme_group",
  data       = df_no_na,
  factorVars = vars_to_show[sapply(df_no_na[vars_to_show], is.factor)]
)
print(tab, quote = FALSE, smd = TRUE)

```
